Outputs from time slice analyses

### 2017-11-28 analyses/02-time-slice-analyses.Rmd

```{r}
## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

@@

## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

>>>>

```{r}
## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
matrix_dist <- readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

@@

## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
matrix_dist <- readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

### 2017-11-28 analyses/03-examine-outputs.Rmd

---
title: "Time slice analyses"
author: "Natalie Cooper and Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

This code runs the analyses in the paper. Below we lay out the process in detail for one dataset, and then provide the code for the other datasets below this. Note that much of this uses functions in the `simulations.R` file.

@@

Make sure you work with `dispRity` 0.4 and TG's version of `Claddis` (much faster than GL's).

```{r, eval = TRUE, message = FALSE}
# The latest version of devtools and ape
library(devtools)
library(ape)

## The latest version of Claddis (on development on TG branch)
## install_github("TGuillerme/Claddis")
library(Claddis)

## The latest version of dispRity on release branch
## install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)

## Load functions using source
# Doesn't always work... Rmd problem?
source("../functions/simulations_fun.R")
source("../functions/tree.age.fossil.R")
```

@@

To run `dispRity` analyses we need the **morphospace**, the **phylogeny** (with a root age and node labels) and the **first and last occurrence dates** (FADLAD). The tree and FADLAD are in the `data/` folder. To construct the morphospace we need to use the `distance_matrix` files we created in `01-extract-data-for analyses.Rmd` and stored in `data/processed`.

```{r}
## Choose a slug
slug <- "Wright2017"

## Loading the discrete morphological matrix
## This is just to match the names with the tree
matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))

## Loading the FADLAD
# FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
```

TG: There's a problem with the FADLAD (too recent! - see 01-extract-data-for-analysis.Rmd lines ~150).
Also, I jus realised that the operations here are done twice, once in the extract-data file and once here.
The clever way would be to do everything in the extract-data (e.g. the root time, rescaling etc...) so that only the time slice analysis is done here.
We could export the data bundle (i.e. the `morphospace`, `tree` and `FADLAD`) in a `.Rda` format and simply load it here (so we're sure we have the correct root time etc.).

```{r}
#TG: Tree tinkering from 01-extract-data-for-analysis.Rmd

# Replace zero branches with minimu branch length
tree$edge.length[which(tree$edge.length == 0)] <- 
  min(tree$edge.length[-which(tree$edge.length == 0)])

# Multiply branches by 0.03517385
tree$edge.length <- tree$edge.length / 0.03517385 #TG: following the note above, it's divide not multiply (i.e. increase branch length because rate < 0)

# Add root time: 485.4
tree$root.time <- 485.4

# Dummy FADLAD
tree_ages <- tree.age.fossil(tree)[1:Ntip(tree),]
FADLAD <- as.data.frame(matrix(c(tree_ages$ages, tree_ages$ages), ncol = 2, byrow = FALSE, dimnames = list(as.character(tree_ages$elements), c("FAD", "LAD"))))

# ## Loading the tree
# tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
# ## Adding node labels and a root age (max tree age date)
# tree <- makeNodeLabel(tree, method = "number", prefix = "n")
# tree$root.time <- 485.4 # specific to Wright 2017 tree

## Cleaning the tree to remove things that aren't in the matrix 
cleaned_data <- clean.data(matrix$matrix, tree)
matrix$matrix <- cleaned_data$matrix
tree <- cleaned_data$tree

## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

## Ordinating the matrix
morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points

Scenarios
--------------------------------

We used the data to explore these different scenarios:

1. Stratigraphy. This is the traditional method, where all the species within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration.

2. Equally sized time bins. 
    i. Where the *duration* of the bin is equal to the median duration of the stratigraphic period.
    ii. Where the *number* of bins is equal to the number of stratigraphic periods.
    iii. Where the *number* of bins is either 2, 5, 10, 15, or 20.

3. Time slicing. 
    i. Where the *interval* between slices is equal to the median duration of the stratigraphic period.
    ii. Where the *number* of slices is equal to the number of stratigraphic periods.
    iii. Where the *number* of slices is either 2, 5, 10, 15, or 20.

Running the analyses and saving the outputs
--------------------------------------------

These analyses use a series of functions in `functions\`. These essentially wrap the existing `dispRity` functions so that running lots of models and extracting lots of data is not such a big issue.

> TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
We now only calculate the disparity for the non-rarefied subsamples and the subsamples rarefied to contain only 3 taxa (the lowest level of rarefaction).
We can use the full rarefaction if we want to plot rarefaction curves later on.

```{r, eval = TRUE, warning = FALSE}
## Ages (rarefied to 3 taxa only)
out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
                                bootstraps = 100, metric = c(sum, variances),
                                metric.name = "sum_var", rarefaction = FALSE)
#readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))

## Epochs (rarefied to 3 taxa only)
out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
                        rarefaction = FALSE)
#readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
```

Note that you can use the previous output format (the data tables) as before by accessing them through `out$results$stratigraphy/duration/number`.
Writing them as a csv will require some thinking since they don't have the same number of columns (though it's still doable I guess).
Also, if you want to plot things through `ggplot`, you can use the values in these data tables!

<!-- TG: I've not run this analysis below -->
<!-- We also ran the models using varying numbers of bins as follows. -->

```{r, eval = FALSE, echo = FALSE}
## Run the function for Beck data
slug <- "BeckLee"

for (bins in c(2, 5, 10, 15, 20)){
  output <- run.all.disparity(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
                              bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")

  readr::write_csv(output, path = paste0("../outputs/bins", "_", slug, ".csv"), append = TRUE)
}

<!-- NC: Need to think more about the outputs here, currently they don't make much sense as there are no column headers -->

@@

>>>>

---
title: "Outputs from time slice analyses"
author: "Natalie Cooper and Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

This code runs the output analyses in the paper, i.e. creates tables, figures and statistical tests.

@@

```{r, eval = TRUE, message = FALSE}
library(devtools)
library(ape)
library(tidyverse)
library(viridis)
library(dispRity) 
```

@@

This script requires the outputs from `02-time-slice-analyses.Rmd`

```{r}
slug <- "Beck2014"
out1 <- readRDS(out1, file = paste0("../outputs/age.", slug, ".Rda"))
out2 <- readRDS(out2, file = paste0("../outputs/age.", slug, ".Rda"))
```

@@

Outputs can be viewed using `out1$results$stratigraphy`, `out1$results$duration`, or `out1$results$number`.
